Think deep for reef resilience

### 2025-09-20 paper/paper_draft.qmd
Remove use of 'Structural' resilience to prevent confusion about temporal resilience and maintenance of carbonate budgets

Understanding the characteristics of reefs that will persist under projected climate
conditions requires examining reef resilience across multiple scales and processes.
Structural resilience - the capacity to maintain coral populations and avoid ecosystem
regime shifts - depends on three interconnected mechanisms. First, the ability of individual
corals to survive thermal stress. Second, the connectivity networks that enable
recolonisation after disturbances [@hock2017; @mumby2008; @torres2024]. Third, the
maintenance of the carbonate framework that provides reef structure [@perry2012]. We aim to
assess how these mechanisms interact under projected climate conditions to identify the
characteristics of structurally resilient reefs through scenario-based coral population
modelling for the Great Barrier Reef ecosystem.

@@

Our findings support the hypothesis that reefs at greater depths will exhibit
characteristics of refugia and more prolonged structural resilience under climate change
compared to shallower reefs. Additionally, Liu, Callaghan and Baldock [-@liu2024] suggest
that rubble instability is inversely correlated with reef depth, possibly limiting juvenile
coral recruitment capacity of shallow reefs. A caveat to our depth-resilience findings is
the quality of the bathymetry data used to calculate median reef depth. As detailed
previously, estimates of median reef depth are derived from satellite imagery (at $10m^2$
resolution) which can be subject to high degrees of uncertainty, both in terms of the
estimated depth for each pixel as well as their alignment to the established GBRMPA reef
outlines. Additionally, we do not incorporate light-based decreases in coral growth rates as
depth increases [@baker1975; @huston1985; @ruiz-diaz2022].

@@

The modelling conducted with ADRIA-CoralBlox itself is subject to many uncertainties. For
one, a formal mechanistic and spatial validation of the model represents an area for future
refinement and is the focus of a future publication. Indicative results are included in the
supplementary **\[TI: For Pedro to include! :`D`\].** The consistency of projected patterns
with known coral-climate relationships and reef characteristics provides confidence in the
model's ability to meaningfully capture ecological signals at the GBR-scale. The mechanistic
basis of the model, combined with qualitative assessment of model behavior, supports the
robustness of our findings for understanding broad-scale coral cover responses to climatic
drivers. The aim here, however, was to assess the relative influence of reef characteristics
on structural resilience under future conditions, rather than provide spatially explicit and
precise predictions of reef outcomes. Using this approach, we suggest that reef depth is an
important factor for reef resilience under thermal stress.

>>>>

Understanding the characteristics of reefs that will persist under projected climate
conditions requires examining reef resilience across multiple scales and processes. Coral
reef ecosystem resilience - the capacity to maintain coral populations and avoid ecosystem
regime shifts - depends on three interconnected mechanisms. First, the ability of individual
corals to survive thermal stress. Second, the connectivity networks that enable
recolonisation after disturbances [@hock2017; @mumby2008; @torres2024]. Third, the
maintenance, and accretion, of the carbonate framework that provides reef structure
[@perry2012]. We aim to assess how these mechanisms interact under projected climate
conditions to identify the characteristics of resilient reefs through scenario-based coral
population modelling for the Great Barrier Reef ecosystem.

@@

Our findings support the hypothesis that reefs at greater depths will exhibit
characteristics of refugia and more prolonged resilience under climate change compared to
shallower reefs. Additionally, Liu, Callaghan and Baldock [-@liu2024] suggest that rubble
instability is inversely correlated with reef depth, possibly limiting juvenile coral
recruitment capacity of shallow reefs. A caveat to our depth-resilience findings is the
quality of the bathymetry data used to calculate median reef depth. As detailed previously,
estimates of median reef depth are derived from satellite imagery (at $10m^2$ resolution)
which can be subject to high degrees of uncertainty, both in terms of the estimated depth
for each pixel as well as their alignment to the established GBRMPA reef outlines.
Additionally, we do not incorporate light-based decreases in coral growth rates as depth
increases [@baker1975; @huston1985; @ruiz-diaz2022].

@@

The modelling conducted with ADRIA-CoralBlox itself is subject to many uncertainties. For
one, a formal mechanistic and spatial validation of the model represents an area for future
refinement and is the focus of a future publication. Indicative results are included in the
supplementary **\[TI: For Pedro to include! :`D`\].** The consistency of projected patterns
with known coral-climate relationships and reef characteristics provides confidence in the
model's ability to meaningfully capture ecological signals at the GBR-scale. The mechanistic
basis of the model, combined with qualitative assessment of model behavior, supports the
robustness of our findings for understanding broad-scale coral cover responses to climatic
drivers. The aim here, however, was to assess the relative influence of reef characteristics
on reef resilience under future conditions, rather than provide spatially explicit and
precise predictions of reef outcomes. Using this approach, we suggest that reef depth is an
important factor for reef resilience under thermal stress.
